Histogram of Gradient Orientations of Signal Plots applied to P300 Detection

### 2017-12-19 response.letter.tex

\documentclass[journal,onecolumn,12pt]{IEEEtran}

\usepackage{amsmath,amssymb,bm}
\usepackage{amsthm, amsfonts}	
\usepackage{bm,bbm}
\usepackage[normalem]{ulem}
\usepackage{color}
\usepackage{fancybox}
\usepackage{url,booktabs}
\usepackage[round]{natbib}

\usepackage{xr}

\title{Reply to Reviewer's Comments on\\
``EEG characterization and classification based on image gradient histograms''\\
(now ``Histogram of Gradient Orientations of Signal Plots applied to P300 Detection'')}
\author{}

@@

\begin{document}

\maketitle
\pagenumbering{roman}
\setcounter{page}{1}

We are grateful to the reviewer for pointing out relevant issues in our manuscript.

@@

2.1) the title is misleading since a first, and naive, read suggest being able to classify and characterize any EEG with their different components (i.e., P200, P300, P600, N100, N200, N400 to name a few). Clearly, the title does not match the content since the authors focus on the P300

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. The title of the article was changed accordingly to your suggestion.  
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

2.2) there is evidence of the benefits of working with these signals P300, but, is the proposed method extensible to other signals?, this is not clear from the manuscript.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
You're  right, thank you very much for your comment.  We published in ~\citep{Ramele2016} the application of the same method to identify rhythmic EEG events like Visual Occipital Alpha Waves and to classify Motor Imagery.  We are also working on unpublished material where we are analyzing the same approach for the detection of K-Complexes and classification of  SSVEP patterns.  We added this information on the article (Introduction and Conclusion, section 3).  We hope this new information enriches the article and clarifies this point.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex
\subsection*{\ovalbox{Third Paragraph}}

3.1) Another issue with the manuscript is its mathematical vagueness. For example, the variable T is used as a vector or two-tuple but its notation is different from the notation the authors use for vectors

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We worked hard to improve all the mathematical description of the method. The issue with variable T was fixed and we improved on the font styling to differentiate vectors and scalars.  Section 1.1(Feature Extraction...) was modified as well as section 1.2.3(Parameters).  The mathematical terminology were corrected all along the article.}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

3.2) Moreover, there is something wrong in equation (4)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We appreciate a lot your truthful comment.  Equation (4) is very important for us because it is the essence of the method.  %We based that equation on the description of how the histogram of oriented gradients is calculated published in Vedaldi and Szeliski, Computer Vision and Algorithms.  
We reviewed the equation, clarified their terms and we hope it is now more clear.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3.3) and something inconsistent between equations (7) and (8)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have revised the manuscript and modified equations (7) and (8) (now (8) and (9)) which are located on Section 1.2.3 (Parameters).  We added more information on the text and specified subscript indexes to avoid the inconsistency.
We thank you for noticing this.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

3.4) Also, the authors go back and forth the digital space and the continuous space and although there is an equivalence, the manuscript could benefit from a clearer explanation.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation.  We modified all equations to emphasize the point that they are calculated in digital space, particularly on equation (4). We also added that image gradients are also calculated using finite differences, and added information about the support of each equation to clarifies that they are actually calculated in digital space.
With this, we hope we have addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

4.1) Another issue that I find worrisome is the design of experiments.\\

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for the observation. 
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

4.2) it seems that the number of subjects is rather low, it would be interesting to know why the authors decided to use such a subset.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your suggestion. The original dataset was published as the 004-2014 dataset at the BNCI-Horizon website, which is very popular in BCI Research.  As far as we are aware of, experiments on this dataset of ALS patients has not been published by any independent group.
In order to address this issue, we included a new own dataset with the same experimental conditions that we generated ourselves, on healthy subjects. We believe it now improves the number of different subjects and that can help to validate the method.  We are also publishing this dataset on the kaggle platform for replication.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

4.3) Also, the authors speculate that their results are not that good because they did not take into account the latency and magnitude of the signals, there is previous evidence in the literature that the P300 suffers from these, and other, problems and, thus, one wonders why the authors did not anticipate such problems.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We added more detailed information that address this issue. First in Section 1.1.4 (Speller Matrix Letter Identification) we included information about how we approach the signal averaging procedure.  In Section 1.2.3 (Parameters) we added the parameters that were used.  More importantly, in Section 2(Results and Discussion) we added information and references about the tests that we did to address latency and amplitude issues.  We included in Section 3(Conclusion) information about future and current works where we are also dealing with these problems.
We hope this is acceptable.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

\subsection*{\ovalbox{Fifth Paragraph}}

5.1) it seems the authors assume that the readers have some previous deep knowledge of some topics such as that some channels should carry more or better information without a previous reference.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We apologize for our mistake. The Results section was modified to clarify more precisely the expectation about the EEG channels where a visual P300 oddball paradigm generates stronger signals.  We added references and improved the section to clarify this point more thoughtfully.
We hope we may have addressed your point with this.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

5.2)  issue has to do with the way the manuscript is written

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you a lot for this important comment.  We re-read all the manuscript and asked  colleagues to proofread the article.  They gave us interesting suggestions that helped us to improve different sections.  We would be very pleased if you find it now easier to read it.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

5.3) sometimes is very difficult to follow the ideas the authors want to convey, an example can be found in the paragraph containing equation (5)

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for this information.  We reviewed completely that equation (it is now equation (6) and equation (7)) and we included a step-by-step procedure that describes the algorithm that we propose and we used as well as a schematic chart in Figure 3.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

\bibliography{response.letter}

### 2017-12-19 article.tex
Revisé varios puntos, reescribí partes, agregué más referencias y completé los Resultados y la Discusión.  Queda pendiente las conclusiones finales

BCI has gained mainstream public awareness with worldwide challenge competitions like Cybathlon~\citep{Riener2014} and even been broadcasted during the inauguration ceremony of the 2014 Soccer World Cup.  New developments have overcome the out-of-the-lab high-bar and they are starting to be used in real world environments~\citep{Huggins2016}.  However, they still lack the necessary robustness, and its performance is well behind any other method of human computer interaction, including any kind of detection of residual muscular movement~\citep{Clerc}.

Several works have explored the idea of exploiting the signal waveform to analyze the EEG signal.  In~\citep{Alvarado-Gonzalez2016} an approach based on Slope Horizontal Chain Code is presented, whereas in ~\citep{Yamaguchi2009} a similar procedure was implemented based on Mathematical Morphological Analysis.  The seminal work of Bandt-Pompe Permutation Entropy~\citep{Berger2017} also explores succinctly this idea as a basis to establish the time series ordinal patterns.  In the article~\citep{Ramele2016},  the authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves  and Motor Imagery rolandic central $\mu$ rhythms using the histogram of gradient orientations of signal plots.  Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  
The proposed approach is based on the waveform analysis of the shape of the EEG signal, but using histogram of gradient orientations. The method is built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots, as it is described in~\citep{Hartman2005}.

@@

Averaged signal segments are scaled and standardized by

@@

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t \in \{0, 1, \dots, t_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to time and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute  
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c), t \in \{0, 1, \dots, t_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

@@

\end{enumerate}
By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $. 
Figure~\ref{fig:classification} shows a schema of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments $S_i$ are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

@@

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  All healthy subjects have normal or corrected-to normal vision and no history of neurological disorders. The experiment is performed with 20 subjects, 6 males, 2 females, 6 right-handed, 2 left-handed, average age 28.25 years, standard deviation  9.58 years, range 20-50 years.

@@

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ Fs $ is the sample frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also provides that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ on the vertical direction and $\frac{1}{F_s.\gamma}=\frac{1}{64}$ seconds on the horizontal direction. Descriptors  $\mathbf{kp}$  are located at $ (x_{kp}, y_{kp} )= ( 0.55 Fs.\gamma, z(c) )= (35,  z(c)) $ (see Eq.~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:sampledescriptor2} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction).

@@

Table~\ref{tab:results} shows the results of applying the algorithm to the subjects of the public dataset of ALS patients. The percentage of correctly spelled letters is calculated while performing an offline BCI Simulation.  From the seven words for each subject, the first three are used as training, and the remaining four for testing.  The best performing channel is informed as well. The chance level is $2\%$.  It can be observed that the best performance of the letter identification method is reached in various channels depending on the subject in study.

@@

In Table~\ref{tab:resultsowndataset} results obtained for the 8 healthy subjects are shown.   It shows the results obtained by the present method as described in previous sections.

@@

The Information Transfer Rate (ITR), or Bit Transfer Rate (BTR), in the case of reactive BCIs~\citep{WolpawJonathanR2012}  depends on the amount of signal averaging required to transmit a valid and robust selection.  Fig.~\ref{fig:performance} shows the performance curves. It can be observed that the percentage of correctly identified letters depends on the number of intensification sequences $k_a$ that are used to obtain the averaged signal.

@@

%\begin{itemize}
%\item PODRIA ACA MOSTRAR ADEMAS PLOTS DE SINGLE TRIAL PARA QUE VEAN QUE SON UN QUILOMBO.%JG no sé, no lo pidió
%\item UNA DE LAS RAZONES POR LA CUAL PODEMOS LLEGAR A ARGUMENTAR QUE NO FUIMOS MAS A FONDO CON EL TEMA ES QUE NO QUERIAMOS DISTORSIONAR DEMASIADO LA SENAL COMPLEJIZANDO EL ALGORITMO Y QUERIAMOS MANTENERNOS ENTRE LO MAS ESTANDAR POSIBLE (PUEDO CITAR MAS O MENOS QUE NUESTRO PROCESSING PIPELINE ES ESTANDARD).%%JG No, a menos que lo pregunte. Tenemos que transmitir que fuimos a fondo en el tema
%\item ACA EL PUNTO PARA MOSTRAR ES QUE DE ALGUNA MANERA ESTAMOS COMO MOSTRANDO UN "SI SE PUEDE" DE UN TRABAJO QUE PUEDE CONTINUAR Y EXPANDIRSE.   POR ESO LA IDEA DE MOSTRAR PUNTAS DE COSAS QUE VIMOS E HICIMOS, PERO SIN MOSTRAR RESULTADOS CONCRETOS.
%\item La pregunta:  ¿ Es posible utilizando solamente la información de los plots, aún utilizando preprocesamiento y una mejora en el SNR basado en el esquema tradicional de generación del patrón de p300, detectarlo con alto grado de éxito ?  %%JG esto puede ir a las conclusiones.

%\begin{itemize}
%\item Es posible pero sólo en algunos pacientes.
%\item Se requiere de mecanismos para estabilizar el waveform de p300.

%\item 
%\end{itemize}
%%JG No entiendo nada
%We find that only by using the visual aspect of the P300 signal plot,  for some subjects it was not possible to find templates that allow  to reach a higher level classification.  
As can be seen in the figure, as the number of intensification sequences tend to 1, which corresponds to single trial letter identification, the performance is reduced. As mentioned before, the SNR of the single trial P300 is very low. The ensemble average is carried out in order to improve the SNR but other problems appear, for example inter-trial variability due to latencies shifts that may produce an unstable P300 signature. To solve this problem, we tested an approach to assess if the morphological shape of the P300 can be stabilized by applying different shifts and we verified that there is a better performance when a correct resynchronization is applied, but results are still in process.  We also applied DTW~\citep{Casarotto2005} but we were unable to find a substantial improvement.     On the other hand amplitude width and latency are generally affected by habituation, fatigue or level of attention and may lead to null signals\citep{Ouyang2017}. This is another source of instability of the P300 signature component that may need to be considered.

@@

Another problem is the amplitude variation of the P300. We propose an approach by standardizing the signal, shown in Eq.~\ref{standarizedaverages}.  It first has the effect of normalizing the peak-to-peak amplitude, moderating its variation. Additionally,  it has the advantage of reducing noise that were not reduced by the averaging procedure.   In addition, the signal variance depends on the number of single-trials used to compute it \citep{van2006signal}.  The standardizing process converts the signal to unit signal variance which makes it independent of the number $k_a$ of signals averaged.   This is another advantage of this process.  On the downsize, the standardizing process reduces the amplitude of any significant P300 complex diminishing its automatic interpretation capability.

For both datasets, the experimental protocol uses a very short inter-stimulus interval which has the potential to increase the ITR but at the same time it reduces the amplitude of the P300 response, hence it may be more difficult to detect~\citep{Rao2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Discussion}
%
%This method, different from other methods which is based on the nonlinearity of the gradient of histograms which can be used to detect 
%
%
%is also based on how the image look like.
%
%SNR of p300 and how to detect it
%
%Check if you can use this to detect any kind of transient signal.
%
%Compare if it is possible with the descriptors from one subject, discriminate the others.
%
%Channal identification based on the metric distance between the bags
%
%Signal Averaging.

@@

%QUE HACER

% Escribir conclusiones buenas sobre todo los mismos puntos que se tocan en la discusion
% Si bien

% El tema del estudio lo mismo que hace el medico pero de manera automatico.

% Letter.  Esto esta aca.

%The SIFT method is invariant to the scale, so in order to allow our method to be also invariant to the amplitude of the signal we had to modify the geometry of the patch to remove the squared constrain.

@@

A method to characterize and classify EEG signals where their characterization is transient in time-space, like the P300 ERP, is presented.

@@

This is due to the fact that the descriptors are directly based on how the signals behave in shape domain (i.e. \textit{how actually they looked like}) for the training and calibration step, and they do not require any prior knowledge about the signal.   In contrast, when the shape of the ERP response is not consistent across the same subject, this method would not be able to find the templates and speller performance could be penalized.

At the same time, by analyzing the generated descriptors, which map in a very detailed and synthetic structure the shape information contained within the patch, a metric about the consistency of the shape of the generated P300 response could also be derived.  It may be worthy of further consideration to evaluate if there is any correlation between the lower performance obtained for some subjects and the characterization of their ALS stage.

Besides novel applications of BCI, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.  It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.

@@

>>>>

BCI has gained mainstream public awareness with worldwide challenge competitions like Cybathlon~\citep{Riener2014} and even been broadcasted during the inauguration ceremony of the 2014 Soccer World Cup.  New developments have overcome the out-of-the-lab high-bar and they are starting to be used in real world environments~\citep{Guger2017,Huggins2016}.  However, they still lack the necessary robustness, and its performance is well behind any other method of human computer interaction, including any kind of detection of residual muscular movement~\citep{Clerc}.

A few works have explored the idea of exploiting the signal waveform to analyze the EEG signal.  In~\citep{Alvarado-Gonzalez2016} an approach based on Slope Horizontal Chain Code is presented, whereas in ~\citep{Yamaguchi2009} a similar procedure was implemented based on Mathematical Morphological Analysis.  The seminal work of Bandt-Pompe Permutation Entropy~\citep{Berger2017} also explores succinctly this idea as a basis to establish the time series ordinal patterns.  In the article~\citep{Ramele2016},  the authors introduce a method for classification of rhythmic EEG events like Visual Occipital Alpha Waves  and Motor Imagery rolandic central $\mu$ rhythms using the histogram of gradient orientations of signal plots.  Inspired in that work, we proposed a novel application of the developed method to classify and describe transient events, particularly the P300 Event Related Potential.  
The proposed approach is based on the waveform analysis of the shape of the EEG signal, but using histogram of gradient orientations. The method is built by mimicking what traditionally electroencephalographers have been performing for almost a century: visually inspecting raw signal plots, as it is described in~\citep{Hartman2005}.

@@

Averaged signal segments are standardized and scaled by

@@

\begin{enumerate}
\item \label{paso1}Highlight randomly the rows and columns from the matrix.  There is one row and one column that should match the letter selected by the subject.
\item  \label{paso2} Repeat step~\ref{paso1} $k_a$ times, obtaining the single trial segments $S_1(t,c),\dots,S_{k_a}(t,c)$, where the variables $t \in \{0, 1, \dots, t_{max}\}$ and $c \in \{1,2,\dots,Ch\}$ correspond to time and channel, respectively. The parameter $k_a$ is the number of repetitions and it is an input parameter of the algorithm.
\item \label{paso3} Compute the Ensemble Average by
\begin{equation}
x(t,c)= \frac{1}{k_a}\sum_{i=1}^{k_a}S_i(t,c), t \in \{0, 1, \dots, t_{max}\}, c \in \{1,\dots,Ch\}
\label{averaging}
\end{equation}  
for each row and column.

@@

\end{enumerate}
By computing the aforementioned equations, the letter of the matrix can be determined from the intersection of the row $ \hat{row} $ and column $ \hat{col} $. 
Figure~\ref{fig:classification} shows a scheme of this process. 
\begin{figure}[htp]
\centering
\includegraphics[width=15cm]{classificationgraph.pdf}
\caption{Single trial segments $S_i$ are averaged for the 6 rows and 6 columns. From the averaged signal, the image of the signal plot is generated and each descriptor is computed.  By comparing each descriptor against the set of templates, the P300 ERP can be detected, and finally the desired letter from the matrix can be inferred.}
\label{fig:classification}
\end{figure}

@@

Participants are recruited voluntarily and the experiment is conducted anonymously in accordance with the declaration of Helsinki published by the World Health Organization.  No monetary compensation is handed out and all participants agree and sign a written informed consent.  All healthy subjects have normal or corrected-to normal vision and no history of neurological disorders. The experiment is performed with 8 subjects, 6 males, 2 females, 6 right-handed, 2 left-handed, average age 28.25 years, standard deviation  9.58 years, range 20-50 years.

@@

\noindent where $ \lambda $ is the length in seconds covered by the patch, $ Fs $ is the sampling frequency of the EEG signal (downsampled to 16 Hz) and  $\Delta  \mu V $ corresponds to the amplitude in microvolts that can be covered by the height of the patch. The geometric structure of the patch forces a squared configuration, then we discerned that by using $ s =s_x =s_y = 3 $ and $ \gamma = 4 $  the local patch and the descriptor can identify events of 9 $ \mu V $ of amplitude, with a span of $ \lambda = 0.56$ seconds.  This also provides that $ 1 $ pixel represents $ \frac{1}{\gamma}= \frac{1}{4} \mu V $ on the vertical direction and $\frac{1}{F_s.\gamma}=\frac{1}{64}$ seconds on the horizontal direction. Descriptors  $\mathbf{kp}$  are located at $ (x_{kp}, y_{kp} )= ( 0.55 Fs.\gamma, z(c) )= (35,  z(c)) $ (see Eq.~\ref{eq:zerolevel}).   In this way the whole transient event is captured. 
Figure~\ref{fig:sampledescriptor2} shows a patch of a signal plot covering the complete amplitude (vertical direction) and the complete span of the signal event (horizontal direction).

@@

Table~\ref{tab:results} shows the results of applying the algorithm to the subjects of the public dataset of ALS patients. The percentage of correctly spelled letters is calculated while performing an offline BCI Simulation.  From the seven words for each subject, the first three are used as training, and the remaining four for testing.  The best performing channel is informed as well. The chance level is $2\%$.  It can be observed that the best performance of the letter identification method is reached in various channels depending on the subject on study.

@@

In Table~\ref{tab:resultsowndataset} results obtained for 8 healthy subjects are shown.   These results were obtained by the present method as described in previous sections.

@@

The Information Transfer Rate (ITR), or Bit Transfer Rate (BTR), in the case of reactive BCIs~\citep{WolpawJonathanR2012}  depends on the amount of signal averaging required to transmit a valid and robust selection.  Fig.~\ref{fig:performance} shows the performance curves for varying intensification sequences. It can be observed that the percentage of correctly identified letters depends on the number of intensification sequences $k_a$ that are used to obtain the averaged signal.

@@

As can be seen in the figure, as the number of intensification sequences tend to 1, which corresponds to single trial letter identification, the performance is reduced. As mentioned before, the SNR of the single trial P300 is very low. The ensemble average is carried out in order to improve the SNR but other problems arise, for example inter-trial variability due to latencies shifts that may produce an unstable P300 signature. To solve this problem, we tested an approach to assess if the morphological shape of the P300 can be stabilized by applying different shifts and we verified that there is a better performance when a correct resynchronization is applied, but results are still in process.  We also applied DTW~\citep{Casarotto2005} but we were unable to find a substantial improvement.     On the other hand amplitude, width and latency are generally affected by habituation, fatigue or level of attention and may lead to null signals\citep{Ouyang2017}. This is another source of instability of the P300 signature component that may need to be addressed.

@@

Another problem is the amplitude variation of the P300. We propose an approach by standardizing the signal, shown in Eq.~\ref{eq:standarizedaverages}.  First, it has the effect of normalizing the peak-to-peak amplitude, moderating its variation. It has also the advantage of reducing noise that were not reduced by the averaging procedure.   It is important to remark that the signal variance depends on the number of single-trials used to compute it \citep{van2006signal}.  The standardizing process converts the signal to unit signal variance which makes it independent of the number $k_a$ of signals averaged.   This is another advantage of this approach.  On the downsize, the standardizing process reduces the amplitude of any significant P300 complex diminishing its automatic interpretation capability.

For both datasets, the experimental protocol uses a very short inter-stimulus interval which has the potential to increase the ITR but at the same time it reduces the amplitude of the P300 response, hence it may be more difficult to detect it~\citep{Rao2013}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Discussion}
%
%This method, different from other methods which is based on the nonlinearity of the gradient of histograms which can be used to detect 
%
%
%is also based on how the image look like.
%
%SNR of p300 and how to detect it
%
%Check if you can use this to detect any kind of transient signal.
%
%Compare if it is possible with the descriptors from one subject, discriminate the others.  Transfer learning.
%
%Channal identification based on the metric distance between the bags
%
%Signal Averaging.

@@

A method to detect transient P300 components from EEG signals based on their characterization in time-space is presented.

Stability of the P300 in terms of shape: Synchronization averaging, increase inter-stimulus interval, different montages, spatial filters.
Collaboration
It can work adaptitevely and also from a set of preconcepts.  This can be used to build a database of descriptors and improve atlases (REFERENCE ATLAS OF EEG).

The SIFT method is invariant to the scale, so in order to allow our method to be also invariant to the amplitude of the signal we had to modify the geometry of the patch to remove the squared constrain.  This will allow us to not use the zscore anymore.

@@

@@

This is due to the fact that the descriptors are directly based on how the signals behave in shape domain (i.e. \textit{how actually they looked like}) for the training and calibration step, and they do not require any prior knowledge about the signal.

In contrast, when the shape of the ERP response is not consistent across the same subject, this method would not be able to find the templates and speller performance could be penalized.

At the same time, by analyzing the generated descriptors, which map in a very detailed and synthetic structure the shape information contained within the patch, a metric about the consistency of the \textit{shape} of the generated P300 response could also be derived.

It may be worthy of further consideration to evaluate if there is any correlation between the lower performance obtained for some subjects and the characterization of their ALS stage.

Besides novel applications of BCI, the goal of the entire discipline is to provide communication assistance to people affected by neuro-degenerative diseases.  It is understood that for people suffering from ALS, even though this pathology do not harm the CNS, their brain signals do suffer slow and deteriorating changes that may lead to different EEG patterns alterations \citep{Nijboer2009,Riener2014}.  Althought this, we did not find such an evidence in terms of the shape.  The best response was obtained by a ALS subject.

@@

Finally, waveform analysis and the methods that can be used to analyze them is an area where active research can benefit greatly both, physicians who need to automate some of the tools and practices and Brain Computer Interfaces or automatic EEG processing by using the extensive knowledge body available for clinical practices.  APUNTAR AL PAPER DE LOTTE MUY BUENO DEL JOURNAL DE BCI.
